---
title: "Spatial Analysis in R, Part 2"
author: "John Fay"
date: "March 21, 2019"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: "cosmo"
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---
# LESSON OBJECTIVES
1. Review basics of vector data analysis
2. Introduce Raster data model
3. Explor raster analysis with `raster` package

A. Import shapefiles: NC Counties and NC HUC8s
B. Import NLCD and DEM data
C. Explore 

## ABOUT RASTER DATA
https://datacarpentry.org/organization-geospatial/01-intro-raster-data/index.html

Raster data is any pixelated (or gridded) data where each pixel is associated with a specific geographical location. The value of a pixel can be continuous (e.g. elevation) or categorical (e.g. land use). If this sounds familiar, it is because this data structure is very common: it’s how we represent any digital image. A geospatial raster is only different from a digital photo in that it is accompanied by spatial information that connects the data to a particular location. This includes the raster’s extent and cell size, the number of rows and columns, and its coordinate reference system (or CRS).

#### Some examples of CONTINUOUS raster data sets:
1. Precipitation maps
2. Maps of tree height derived from LiDAR data
3. Elevation values for a region 

#### Some examples of CATEGORIGAL raster data sets:
1. Landcover / land-use maps.
2. Tree height maps classified as short, medium, and tall trees.
3. Elevation maps classified as low, medium, and high elevation.

#### Attributes of raster data sets:
1. Extent
2. Cell size/resolution
3. Coordinate reference system
4. Values representing missing data

#### Multi-band raster data: 
1. Visual color bands: RGB
2. Multispectral/hyper spectral imagery

#### Raster data files and formats
1. ASCII (TIFF, BIN, BIN) vs Binary (ASCII)
2. Headers and World files

```{r Prepare for analysis}
getwd()
#Old libraries
library(tidyverse)
library(sf)
library(mapview)
library(leaflet)
library(RColorBrewer)

#New ones
#import.packages('raster')
library(raster)
#import.packages('FedData')  #<-- useful for extracting spatial data
library(FedData)
```


```{r Import and transform HUC and County Data to a common CRS}
#Read in the HUCs and save its CRS to a variable
nc_hucs <- st_read('./Data/Spatial/huc_250k_nc.shp')
st_crs(nc_hucs)
plot(nc_hucs['SUB'])

#Read in the counties, filtering for Durham, Orange, Chatham, and Wake
tri_counties <- st_read('./Data/Spatial/cb_2017_us_county_20m.shp') %>% 
  filter(STATEFP == 37 & NAME %in% c('Durham', 'Orange', 'Chatham', 'Wake'))
st_crs(tri_counties)

#Transform the counties sf to match the hucs sf
tri_counties <- st_transform(tri_counties,st_crs(nc_hucs))
st_crs(tri_counties)

#Select the HUCs that intersect the counties
hucMask <- st_intersects(nc_hucs,st_union(tri_counties),sparse = FALSE)
tri_hucs <- nc_hucs[hucMask,]

#Plot the results
ggplot() + 
  geom_sf(data=nc_hucs,aes(fill = SUB), color='white') +   #Show all HUC8s
  geom_sf(data=tri_hucs, fill = NA, size = 1) +            #Show selected HUC8s
  geom_sf(data=tri_counties, fill=NA, color='black')       #Add the counties
```


Fetch elevation and land cover data using the FedData package's `get_nlcd` command:
```{r Fetch NLCD data, eval=FALSE}
#Fetch the data
nlcd2011 <- get_nlcd(template = tri_counties,             #Template specifies the extent to fetch
                     label='tri',                         #Sets a label for the output
                     year=2011,                           #The year of the land cover dataset to fetch
                     dataset='landcover',                 #The dataset to fetch
                     raw.dir='./Data/Spatial/tmpNLCD',    #Where to fetch the tiles (creates this folder)
                     extraction.dir = './Data/Spatial')   #Where to store the merged tiles (final result)
```

```{r}
#If the above fails or takes too long, uncomment and use the command below
#nlcd2011 <- raster('./Data/Spatial/tri1_NLCD_2011_landcover.tif')

#Plot the data
plot(nlcd2011)

#Uncomment to remove all temporary tiles
#unlink('./Data/Spatial/tmpNLCD',recursive = TRUE)
```
## IMPORTANT: These downloads may exceed your GitHub push limit. Before pushing, delete all intermediate data - nothing over 100MB!! 

Now you try it: Fetch elevation data from the National Elevation Dataset (NED). The command is `get_ned` and its parameters are fewer than NLCD: 
* use the same sf dataframe as your `template` (the `tri_counties` dataframe)
* use the same `label`, `raw.dir`, and `extraction.dir` as above too. 
```{r Fetch Elevation data, eval=FALSE}
#Fetch elevation data for the "tri_counties" extent into a raster called "dem30"
dem30 <- get_ned(template = tri_counties,
                 label='tri',
                 raw.dir='./Data/Spatial',
                 extraction.dir = './Data/Spatial')
```

```{r Resample the 30m elevation data to 90 meters}
#Resample the 30m data to 90m (factor of 3)
dem90 <- aggregate(dem30,fact=3,fun=mean,filename='./Data/Spatial/tri_NED_90.tif')

#If the NED extraction & aggergation fails, uncomment and use the dataset below
#dem90 <- raster('./Data/Spatial/tri1_NED_90.tif')

#Plot the data
plot(dem90)


#Uncomment and run to delete the 30m tiles 
#unlink('./Data/Spatial/1', recursive = TRUE)
#remove.file('tri_NED_1.tif')

```

## Raster data
A raster is fundamentally a data matrix, and individual pixel values can be
extracted by regular matrix subscripting. For example, the value of
the _bottom_-left corner pixel:

```{r get_raster_values}
nlcd2011[1, 1]
```


The meaning of this number is not immediately clear. For this particular
dataset, the mapping of values to land cover classes is described in the data
attributes:

```{r raster_attr}
head(nlcd2011@data@attributes[[1]])
```


## Visualization options
```{r}
#View a histogram of the pixel values
hist(dem90)

#View with the image command
image(dem90,col = terrain.colors(20))

#View a slice of data a different color
image(dem90, zlim=c(50,80), add=TRUE)

#Add contours
contour(dem90,add=TRUE)

```


## Raster math
```{r}
dem_ft = dem90 * 3.28
dem_ft
plot(dem_ft)
```


## Raster tranformations
Raster projections require Pro4strings, which you can get from the spatialreference.org web site. Here we get the proj4 string for UTM Zone 17N (epsg = 26917)
```{r}
dem90_utm <- projectRaster(dem90,crs='+proj=utm +zone=17 +ellps=GRS80 +datum=NAD83 +units=m +no_defs')
image(dem90_utm)
```


## Raster functions
```{r}
slope90 <- terrain(dem90_utm,opt='slope')
hist(slope90)
image(slope90,zlim= c(0,0.1),col = topo.colors(45))
```
> What are some other terrain functions? 


## Cropping and masking raster rasters
```{r}
triNLCD_cropped <- crop(nlcd2011, tri_counties %>% filter(NAME=='Durham'))
plot(triNLCD_cropped)

triNLCD_masked <- mask(triNLCD_cropped, tri_counties %>% filter(NAME=='Durham'))
plot(triNLCD_masked)

hist(triNLCD_masked)

urban <- mask(triNLCD_cropped, triNLCD_cropped %in% c(21,22,23,24), maskvalue = FALSE)
plot(urban)
```

