---
title: "Assignment 8: Time Series Analysis"
author: "Yifei Zhang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on time series analysis.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A08_TimeSeries.pdf") prior to submission.

The completed exercise is due on Tuesday, 19 March, 2019 before class begins.

## Brainstorm a project topic
1. Spend 15 minutes brainstorming ideas for a project topic, and look for a dataset if you are choosing your own rather than using a class dataset. Remember your topic choices are due by the end of March, and you should post your choice ASAP to the forum on Sakai.

Question: Did you do this?

> ANSWER: Yes.

## Set up your session 
2. Set up your session. Upload the EPA air quality raw dataset for PM2.5 in 2018, and the processed NTL-LTER dataset for nutrients in Peter and Paul lakes. Build a ggplot theme and set it as your default theme. Make sure date variables are set to a date format.

```{r}
getwd()
library(tidyverse)
library(nlme)
library(trend)
EPAair <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
PeterPaul <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")

yifeitheme <- theme_light(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(yifeitheme)

PeterPaul$sampledate <- as.Date(PeterPaul$sampledate, format = "%Y-%m-%d")
EPAair$Date <- as.Date(EPAair$Date, format = "%m/%d/%y")

```


## Run a hierarchical (mixed-effects) model

Research question: Do PM2.5 concentrations have a significant trend in 2018?

3. Run a repeated measures ANOVA, with PM2.5 concentrations as the response, Date as a fixed effect, and Site.Name as a random effect. This will allow us to extrapolate PM2.5 concentrations across North Carolina.

3a. Illustrate PM2.5 concentrations by date. Do not split aesthetics by site.

```{r, warning = FALSE}
ggplot(EPAair, aes(x = Date, y = Daily.Mean.PM2.5.Concentration))+
  geom_point()+
  labs(x="")+
  ylab(expression(paste("PM2.5 Concentration(\U003BCg/m"^3,")")))
```

3b. Insert the following line of code into your R chunk. This will eliminate duplicate measurements on single dates for each site.
PM2.5 = PM2.5[order(PM2.5[,'Date'],-PM2.5[,'Site.ID']),]
PM2.5 = PM2.5[!duplicated(PM2.5$Date),]

3c. Determine the temporal autocorrelation in your model. 

3d. Run a mixed effects model. 

```{r, warning = FALSE}
EPAair = EPAair[order(EPAair[,'Date'],-EPAair[,'Site.ID']),]
EPAair = EPAair[!duplicated(EPAair$Date),]

PM2.5Test_auto <- lme(data = EPAair,
                      Daily.Mean.PM2.5.Concentration ~ Date, 
                     random = ~1|Site.Name)
PM2.5Test_auto

ACF(PM2.5Test_auto)

PM2.5Test_mixed <- lme(data = EPAair,
                     Daily.Mean.PM2.5.Concentration ~ Date, 
                     random = ~1|Site.Name,
                     #specify autocorrelation structure of order 1
                     correlation = corAR1(form = ~ Date|Site.Name, value = 0.474),
                     #define method as restricted maximum likelihood
                     method = "REML")
summary(PM2.5Test_mixed)
```

Is there a significant increasing or decreasing trend in PM2.5 concentrations in 2018? 

> ANSWER: The p-value is 0.36 > 0.05, which indicates there is not a significant trend in PM2.5 concentration in 2018.

3e. Run a fixed effects model with Date as the only explanatory variable. Then test whether the mixed effects model is a better fit than the fixed effect model. 

```{r, warning = FALSE}
PM2.5Test_fixed <- gls(data = EPAair,
                      Daily.Mean.PM2.5.Concentration ~ Date, 
                      method = "REML")
summary(PM2.5Test_fixed)

anova(PM2.5Test_mixed, PM2.5Test_fixed)
```


Which model is better?

> ANSWER:  The mixed effect model is better due to the lower AIC


## Run a Mann-Kendall test

Research question: Is there a trend in total N surface concentrations in Peter and Paul lakes? 

4. Duplicate the Mann-Kendall test we ran for total P in class, this time with total N for both lakes. Make sure to run a test for changepoints in the datasets (and run a second one if a second change point is likely). 

```{r, warning = FALSE}
PeterPaul.surface <- 
  PeterPaul %>%
  select(-lakeid, -depth_id, -comments) %>%
  filter(depth == 0) %>%
  filter(!is.na(tn_ug))

# Initial visualization of data
ggplot(PeterPaul.surface, aes(x = sampledate, y = tn_ug, color = lakename)) + 
  geom_point() +
  scale_color_manual(values = c("#7fcdbb", "#253494"))

# Split dataset by lake
Peter.surface <- filter(PeterPaul.surface, lakename == "Peter Lake")
Paul.surface <- filter(PeterPaul.surface, lakename == "Paul Lake")

mk.test(Peter.surface$tn_ug)
mk.test(Paul.surface$tn_ug)
#Alternative hypothesis: there is a trend happenning over time

# Test for change point
pettitt.test(Peter.surface$tn_ug)

# Run separate Mann-Kendall for each change point
mk.test(Peter.surface$tn_ug[1:35])
mk.test(Peter.surface$tn_ug[36:98])

# Test for a second change point
pettitt.test(Peter.surface$tn_ug[36:98])

# Run another Mann-Kendall for the second change point
mk.test(Peter.surface$tn_ug[36:56])
mk.test(Peter.surface$tn_ug[57:98])
```


What are the results of this test?

> ANSWER: There is not a trend over time in total N surface concentration in Paul lake.(Mann-Kendall test, p-value = 0.7) For Peter lake, there is a trend over time in total N surface concentrations, and the two change points are at 1993-06-05 and 1994-06-30.(pettitt test, p-value <0.0001)

5. Generate a graph that illustrates the TN concentrations over time, coloring by lake and adding vertical line(s) representing changepoint(s).

```{r, warning = FALSE}
ggplot(PeterPaul.surface, aes(x = sampledate, y = tn_ug, color = lakename)) + 
  geom_point() +
  geom_vline(xintercept = as.Date("1993-06-02"),color="yellow", lty = 2)+
  geom_vline(xintercept = as.Date("1994-06-29"),color="yellow", lty = 2)+
  scale_color_manual(values = c("#7fcdbb", "#253494"))+
  labs(x = "", y = "total N(\U003BCg/L)")

```

